Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-03 Manuscript/manuscript.tex
committing what should have been contained in previous commit.

\begin{abstract}
  % insert abstract here
  Despite of the significant biological relevance, the binding details of cations in cellular
  lipid membranes are not fully understood. For example, consistent results for
  the binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  achieved with different experimental and theoretical methods.
  The ion binding details could be resolved with classical molecular dynamics (MD) simulations,
  however, the accuracy of the available lipid models has not been sufficient.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipid model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipid model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the ternary complex binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared spectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved by implicitly including the electronic polarizability
in the polar region of lipid molecules. The electronic
polarizability is included by using the electronic
continuum correction (ECC)~\cite{leontyev11}, which
has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
As a starting point we use the parameters from the Lipid14 model \cite{dickson14},
which gave the best cation binding behaviour in the previous study \cite{catte16}.
The developed ECC-lipid parameters reproduce the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly for reproducing the membrane binding affinities and induced
structural effects of Sodium and Calcium ions.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered a higly relevant issue since the early days of
lipid bilayer simulations.
In this work we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{lucas12,chowdhary13} by implicitly including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
earlier studies \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids DOI: 10.1002/anie.201308760}
The present concept of ECC is, however, 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11}.

Following the ECC concept, electronic
polarizability can be effectively included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. 
Following the Coulomb's law, such a dielectric continuum can be easily included in standard MD simulation by
a scaling of charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
Given the  high frequency dielectric constant as measured in water (corresponding to the square of the refraction index) of  $\epsilon _{el} = 1.78$,  
the scaling factor for aqueous ions is $f_q = 0.75$ \cite{some_orig_source, leontyev11}. This scaling factor has been
successfully used to improve the performance of force field for ions in solution \cite{kohagen14,kohagen16,??},
which then agree quantitatively with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014}.

While the scaling factor of $f_q = 0.75$ for ions in water improves their description and is physically well justified within the ECC theory \cite{Leontyev2011},
it is not a priori  clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an atom or molecules, atomic partial charges within molecules are not physical observables.
There are thus several schemes for the assignment of partial charges for biomolecules~\cite{Hu2007}.
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
By construction, partial charges currently implemented in force fields may already 
include  to some extend the solvent electronic polarizability effects, i.e.,
the RESP charges are often scaled to fit certain experimental observables \todo{This needs a citation - Joe please pick 1-2 papers}. 
Thus, a consistent application of the ECC scaling factor, $f_q$, to the molecular partial charges
included in the available force fields does not necessarily have to follow the above relation $f_q = \epsilon _{el} ^{-1/2}$, 
but instead it lies between 0.75 i.e., (no electronic polarizability included in the original partial charges)
and $1$ (i.e., electronic polarizability fully included in the original partial charges).
It is important to note that even thought the lipid bilayer interface has a large gradient in
the total dielectric constant, the electronic polarizability part can be considered to be constant
as the measurements of high frequency dielectric constant gives values around 2 
for almost any biologically relevant material \cite{some_original_work, leontyev11}.

In this work, we employ the ECC to develop a classical MD simulation model for
a POPC lipid that accurately describes the binding of Sodium and Calcium ions to a lipid bilayer.
%This is a biologically highly relevant membrane feature, which is
%poorly reproduced by currently available models. This can affect not only on membrane properties
%in the presence of ions but also modulate interactions with charged moieties at the interface.
The model quality in terms of the ion binding is measured by directly
comparing simulations to the experimental NMR data used to measure the ion binding
affinity~\cite{akutsu81,altenbach84,scherer89}, as discussed in detail in Ref. \citenum{catte16}.
The Lipid14~\cite{dickson14} force field parameters (available in Gromacs format from Ref. \citenum{lipid14files})
were used as a starting point, because they give the most
realistic response to the bound ions and to the NaCl concentration 
among the available lipid models (see Figs.~5~and~2 in Ref.~\citenum{catte16}).
Moreover, the Lipid14 model gave relatively realistic glycerol backbone and
head group structures when compared with the other state of the art lipid models~\cite{botan15}.

The ECC correction was applied to the Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone, and carbonyl regions, which are the most polar parts in the  lipids and
are thus expected to have the largest contribution to the cation binding.
We do not modify the hydrocarbon chain parameters, as they
do not come in direct contact with salt ions and are
already highly optimized to provide a good description for the
hydrophobic part of lipid bilayers \cite{dickson14,ollila16}.
This is in contrasts with the glycerol backbone and the head group regions, 
which call for improvements in all available lipid models~\cite{botan15}.

The values within the above discussed parameter space for the scaling factor, $f_q~\in~(0.75, 1.0)$,
were explored to find the optimal value to reproduce the experimental ion binding affinity.
We observed that the scaling
of the partial charges reduced the ion binding affinity and the 
related head group order parameter response in general.
By comparing the results with different scaling factors to the experimental 
NMR data~\cite{akutsu81,altenbach84,scherer89} we found the optimal 
value of $f_q = 0.8$,
which is only slighly higher than the scaling factor for the ions in water 0.75.
%, which implies
%that electronic polarization effects are only to a small extent effectively accounted for in the Lipid14 force field.

While, the scaling of the partial charges improved the ion binding affinity,
it reduced the bilayer area per lipid below the experimental value. 
%Simulations with Lipid14 parameters having 
After scaling the partial charges of the head group, the glycerol
backbone and carbonyls with a factor of 0.8, the simulation gave 
the approximately 60~\AA$^2$ for the area per lipid,
which is somewhat smaller than the experimental value, 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value given by the original Lipid14 parameters, $65.6 \pm 0.5$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region,
due to the lower polarity of molecules with scaled charges. This can be compensated
by reducing the effective radius of atoms with the scaled charges by changing 
the $\sigma$ parameters in the Lennard-Jones potential, 
as previously done for ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing these $\sigma$ parameters by a factor of $f_\sigma = 0.89$, the area per molecule is restored
back close to the experimental value (see Table~\ref{tab:apls}).

\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules. \\
  JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer
  charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever)
  and ``semi-scaled'' models. \\
SAMULI: There is now a paragraph in the conclusions, which mentions this topic.}

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations by 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of charge bound per lipid~\cite{seelig87}.
The change of order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

@@

The relation between the amount of the bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with bound positive charge and increase with bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt toward water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

The measured order parameter change depends on both, the head group response 
to the bound charge and the amount of bound charge, i.e. $m_i$ and $X^\pm$
in Eq.~\ref{OPchangeEQ}, respectively. The former property has to be
well quantified before the electrometer concept can used to 
analyze the binding affinities, as done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response to the bound charge in simulations,
we use the experimental data for dihexadecyldimethylammoniumbromide %(2C$_{16}$$^+$N2C$_1$Br$^-$)
mixed within a POPC bilayer~\cite{scherer89}. Dihexadecyldimethylammonium\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
The early electrometer concept data for the lipid head group
order parameter changes from the NMR experiments was reported as a function 
of the salt concentration in water before solvating
the lipids~\cite{akutsu81}. The later study used also the atomic absorption
spectroscopy and reported the order parameter changes as
a function of the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was calculated from 
the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the ion concentrations were calculated in water before solvating the lipids 
as in the earlier experiments~\cite{akutsu81}.
Despite of the measurable differences between these two
definitions of concentrations for CaCl$_2$ systems,
the qualitative conclusions in this or in the previous work~\cite{catte16} are
not affected by this.

To quantify the ion binding affinity to a lipid bilayer, we
calculated the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
This quantity was chosen because it does not depend on
%the definition of the interfacial region or
the position of the Gibbs dividing plane between two bulk regions.
%\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
%SAMULI: Here is a new trial.
Here we assume that the interface locates between the hydrophobic interior
of a lipid bilayer and the bulk water region far from the membrane.
The bulk concentration of ions and water is zero inside the bilayer.
The concentrations in bulk water region can be calculated from 
the farthest point from both lipid leaflets in the water phase.
The region between these boundaries contains all the ions and water
molecules in the simulation box.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the area of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

The order parameters were calculated from simulations for all C-H bonds
in lipid molecules by using Eq. \ref{OP}. Form factors were calculated 
from equation \todo{As Hector suggested, it might be better to write the simpler
form for this equation.}
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length in the bulk region,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning until the water bulk region, $D$.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
  \todo{Labeling should be consistent with the previous figure, i.e. experimental data with diamonds.} \\
  \todo{x-axis scale from -1 to 51 to make the point in zero fully visible and to remove empty space.} \\
  \todo{Empty space between figures and from the right column could be reduced.} \\
  \todo{Put the experimental data with larger points and thicker line in this and in the next two figures to
    make more clear separation between simulation and experimental data.}
\end{figure}

>>>>

\begin{abstract}
  % insert abstract here
  The binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  consistently determined with different experimental and theoretical methods
  despite their significance and biological relevance in cellular membranes. 
  The ion binding details could be resolved with classical molecular dynamics (MD) simulations,
  however, the accuracy of the available lipid models has not been sufficient.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipid model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipid model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the ternary complex binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared spectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ is observed to specifically
bind to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved with an implicit inclusion of the electronic polarizability
in the polar region of phospholipids,
which we apply according to the 
electronic continuum correction (ECC)~\cite{leontyev11}.
Such an approach has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
We use the the Lipid14 model \cite{dickson14} as a starting point,
because it provided one of the best available descriptions of cation binding \cite{catte16}.
The developed ECC-lipid model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly in the description of the membrane binding affinities to cations and induced
structural effects of sodium and calcium ions.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of polarizability in standard MD simulation
force fields has been considered an issue since the early days of
lipid bilayer simulations.
In this work, we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{lucas12,chowdhary13},
including the electronic part of polarizability in lipid
bilayer simulations implicitly through 
the electronic continuum correction (ECC)~\cite{leontyev11}. 
Technically, it is similar to the phenomenological charge-scaling applied in 
earlier studies \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids DOI: 10.1002/anie.201308760}
However, the present concept of ECC is
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

According to ECC, electronic
polarizability can be implicitly included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. 
Following the Coulomb's law, such a dielectric continuum can be 
easily included in a standard MD simulation by
the scaling of charges,
\begin{equation}
  Q^{ECC} = f_q \cdot Q ,
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
Given the  high frequency dielectric constant as measured in water (corresponding to the square of the refraction index) of  $\epsilon _{el} = 1.78$,  
the scaling factor for aqueous ions is $f_q = 0.75$ \cite{some_orig_source, leontyev11}. This scaling factor has been
successfully used to improve the performance of force field for ions in solution \cite{kohagen14,kohagen16,??},
which then agree quantitatively with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014}.

While the scaling factor of $f_q = 0.75$ for ions in water improves their description and is physically well justified within the ECC theory \cite{Leontyev2011},
it is not a priori  clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an atom or molecules, atomic partial charges within molecules are not physical observables.
There are thus several schemes for the assignment of partial charges for biomolecules~\cite{Hu2007}.
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
By construction, partial charges currently implemented in force fields may already 
include  to some extend the solvent electronic polarizability effects, i.e.,
the RESP charges are subsequently modified to fit certain experimental observables \todo{This needs a citation - Joe please pick 1-2 papers}. 
Thus, a consistent application of the ECC scaling factor, $f_q$, to the molecular partial charges
included in the available force fields does not necessarily have to follow the above relation $f_q = \epsilon _{el} ^{-1/2}$, 
but instead it lies between 0.75 i.e., (no electronic polarizability included in the original partial charges)
and $1$ (i.e., electronic polarizability fully included in the original partial charges).
It is important to note that even though the lipid bilayer interface has a large gradient in
the total dielectric constant, the electronic part of polarizability can be considered to be constant
as the measurements of high frequency dielectric constant gives values around 2 
for almost any biologically relevant material \cite{some_original_work, leontyev11}.

In this work, we employ the ECC to develop a classical MD simulation model for
a POPC lipid that accurately describes the binding of sodium and calcium ions to a lipid bilayer.
%This is a biologically highly relevant membrane feature, which is
%poorly reproduced by currently available models. This can affect not only on membrane properties
%in the presence of ions but also modulate interactions with charged moieties at the interface.
The quality of the simulation model is compared 
to the experimental NMR data used to measure the ion binding
affinity~\cite{akutsu81,altenbach84,scherer89}, 
as discussed in detail in Ref.~\citenum{catte16}.
The Lipid14~\cite{dickson14} force field parameters (available in Gromacs format from Ref. \citenum{lipid14files})
were used as a starting point, because they provide one of the most
realistic descriptions of the head group order parameter response to ions 
among the available lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}).
Moreover, the Lipid14 model has a relatively realistic glycerol backbone and
head group structure compareable to other state of the art lipid models~\cite{botan15}.

We applied the ECC correction to the Lipid14 model of POPC as a scaling of partial charges of the head group,
glycerol backbone, and carbonyl regions, which are the most polar parts in phospholipids
being expected to have the largest contribution to the cation binding.
We do not modify the hydrocarbon chain parameters, 
as they do not come in direct contact with salt ions, 
and as they are already highly optimized to provide a good description of the
hydrophobic part of lipid bilayers \cite{dickson14,ollila16}.
This is in contrasts with the glycerol backbone and the head group regions, 
which call for improvements in all available lipid models~\cite{botan15}.

The values within the above discussed parameter space for the scaling factor, $f_q~\in~(0.75, 1.0)$,
were explored to find the optimal value to reproduce the experimental ion binding affinity.
We observed that the scaling
of the partial charges reduced the ion binding affinity and the 
related head group order parameter response in general.
We compared the results with different scaling factors to the experimental 
NMR data~\cite{akutsu81,altenbach84,scherer89} finding the optimal 
value $f_q = 0.8$,
which is only slighly higher than the scaling factor for the ions in water, 0.75.

Although the scaling of the partial charges improved 
the head group order parameter response and ion binding affinity,
it also deteriorated some membrane properties; 
namely the area per lipid decreased below the experimental value. 
After only scaling the partial charges of the head group, the glycerol
backbone and carbonyls with a factor of 0.8, 
the area per lipid in the simulation drops to approximately 60~\AA$^2$,
which is significantly smaller than the experimental value, 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value from the original Lipid14 model, $65.6 \pm 0.5$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region
as a consequence of the lower polarity after scaling charges. 
This effect can be compensated
by reducing the effective radii of atoms with the scaled charges.
This is in practice done by changing the parameters $\sigma$ in the Lennard-Jones potential 
in a similar way as was previously done for the ECC-ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing the $\sigma$ parameters of affected atoms with a scaling factor $f_\sigma = 0.89$, 
the area per molecule is restored
back close to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group
(Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of charge bound per lipid~\cite{seelig87}.
The change of the order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

@@

The relation between the amount of the bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with a bound positive charge and increase with a bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt towards water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

The measured order parameter change depends on both the head group response 
to the bound charge, and the amount of bound charge 
(i.e. $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively). 
The empirical factor $m_i$ has to be
well quantified before the electrometer concept can used to 
analyze the binding affinities.
This was done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response to the bound charge in simulations,
we use the experimental data for dihexadecyldimethylammoniumbromide %(2C$_{16}$$^+$N2C$_1$Br$^-$)
mixed within a POPC bilayer~\cite{scherer89}. Dihexadecyldimethylammonium\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
The concentrations of salts are reported in two different ways
in the NMR measurements of head group order parameters \cite{akutsu81, altenbach84}.
The early work on the NMR electrometer concept 
reports the salt concentrations in water before solvating
the lipids~\cite{akutsu81}. 
The later study uses the atomic absorption
spectroscopy, and reports the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work, we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was calculated from 
the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study~\cite{catte16},
the ion concentrations were calculated in water before solvating the lipids 
as in the earlier experiments~\cite{akutsu81}.
Although there are measurable differences between these two
definitions of concentrations for CaCl$_2$ systems,
they do not affect the qualitative conclusions in this or in the previous work~\cite{catte16}.

To quantify the ion binding affinity to a lipid bilayer, we
calculate the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
Such a quantity compares the adsorption of ions to the adsorption of water molecules 
at an interface without the necessity of defining the Gibbs dividing plane 
between the membrane interior and the water bulk region. 
In our simulations we assume that the interface locates between the hydrophobic interior
of a lipid bilayer and the bulk water region far from the membrane.
The bulk concentration of ions and water is zero inside the bilayer.
The concentrations in bulk water region can be calculated from 
the farthest point from both lipid leaflets in the water phase.
The region between these boundaries contains all the ions and water
molecules in the simulation box.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
Such a setup provides a simplified relation for $\Gamma_i^w$ for simulations of lipid bilayers, 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the area of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

The order parameters were calculated from simulations for all C-H bonds
in lipid molecules from Eq. \ref{OP}. Form factors were calculated 
from the equation
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \rho_{el}(z) - \rho_{el}^s \right ) \cos (zq_z) \mathrm{d}z,
  %F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $\rho_{el} (z)$ is the total electron density,
$\rho_{el}^s$ is the electron density of the solvent far in the bulk,
and $z$ is the distance from the membrane centre along its normal 
spanning until the water bulk region, $D/2$.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
\end{figure}
